---
title: "Interactive Data Visualization with Plotly"
author: "Lukhanyiso Bvuma"
date: "`r Sys.Date()`"
output:
  slidy_presentation:
    theme: default
    highlight: tango
  ioslides_presentation:
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plotly)
library(dplyr)
```

## Welcome to Interactive Visualizations!

**Date Created:** `r format(Sys.Date(), "%B %d, %Y")`

This presentation showcases the power of Plotly for creating interactive, web-based visualizations in R.

**What makes Plotly special?**

- Hover for details
- Zoom and pan
- Click to filter
- Download plots
- Mobile-friendly

---

## Dataset: mtcars

We'll explore the classic `mtcars` dataset with interactive plots.

```{r data_preview}
knitr::kable(head(mtcars, 5), caption = "Sample of mtcars dataset")
```

**Variables:**
- `mpg`: Miles per gallon
- `hp`: Horsepower  
- `wt`: Weight (1000 lbs)
- `cyl`: Number of cylinders

---

## Interactive Scatter Plot

```{r scatter_plot, fig.width=9, fig.height=6}
plot_ly(mtcars, 
        x = ~wt, 
        y = ~mpg,
        type = "scatter",
        mode = "markers",
        color = ~factor(cyl),
        size = ~hp,
        text = ~paste("Car:", rownames(mtcars),
                     "<br>MPG:", mpg,
                     "<br>Weight:", wt,
                     "<br>HP:", hp,
                     "<br>Cylinders:", cyl),
        hoverinfo = "text",
        marker = list(opacity = 0.7,
                     line = list(width = 1, color = "white"))) %>%
  layout(title = "Fuel Efficiency vs Weight",
         xaxis = list(title = "Weight (1000 lbs)"),
         yaxis = list(title = "Miles per Gallon"),
         legend = list(title = list(text = "Cylinders")))
```

**Try hovering over the points!** Notice the relationship between weight and fuel efficiency.

---

## 3D Interactive Plot

```{r 3d_plot, fig.width=9, fig.height=7}
plot_ly(mtcars, 
        x = ~wt, 
        y = ~hp, 
        z = ~mpg,
        type = "scatter3d",
        mode = "markers",
        color = ~factor(cyl),
        size = ~disp,
        text = ~rownames(mtcars),
        marker = list(opacity = 0.8)) %>%
  layout(title = "3D View: Weight, Horsepower & MPG",
         scene = list(
           xaxis = list(title = "Weight"),
           yaxis = list(title = "Horsepower"),
           zaxis = list(title = "MPG")
         ))
```

**Try rotating the plot!** Click and drag to explore different angles.

---

## Box Plot: Distribution by Cylinders

```{r box_plot, fig.width=9, fig.height=6}
plot_ly(mtcars, 
        x = ~factor(cyl), 
        y = ~mpg,
        type = "box",
        color = ~factor(cyl),
        boxpoints = "all",
        jitter = 0.3,
        pointpos = -1.8,
        text = ~rownames(mtcars),
        hoverinfo = "text+y") %>%
  layout(title = "Fuel Efficiency by Number of Cylinders",
         xaxis = list(title = "Number of Cylinders"),
         yaxis = list(title = "Miles per Gallon"),
         showlegend = FALSE)
```

**Insight:** Fewer cylinders generally mean better fuel efficiency.

---

## Bar Chart: Average MPG by Cylinder

```{r bar_chart, fig.width=9, fig.height=6}
mpg_by_cyl <- mtcars %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg),
            count = n(),
            .groups = "drop")

plot_ly(mpg_by_cyl,
        x = ~factor(cyl),
        y = ~avg_mpg,
        type = "bar",
        text = ~paste("Avg MPG:", round(avg_mpg, 1),
                     "<br>Count:", count),
        hoverinfo = "text",
        marker = list(color = c("#FF6B6B", "#4ECDC4", "#45B7D1"),
                     line = list(width = 1.5, color = "white"))) %>%
  layout(title = "Average Fuel Efficiency by Cylinder Count",
         xaxis = list(title = "Number of Cylinders"),
         yaxis = list(title = "Average MPG"))
```

---

## Correlation Heatmap

```{r heatmap, fig.width=9, fig.height=7}
# Select numeric columns
numeric_cols <- c("mpg", "cyl", "disp", "hp", "drat", "wt", "qsec")
cor_matrix <- cor(mtcars[, numeric_cols])

plot_ly(z = cor_matrix,
        x = colnames(cor_matrix),
        y = rownames(cor_matrix),
        type = "heatmap",
        colorscale = "RdBu",
        zmid = 0,
        text = round(cor_matrix, 2),
        hovertemplate = "X: %{x}<br>Y: %{y}<br>Correlation: %{z:.2f}<extra></extra>") %>%
  layout(title = "Correlation Matrix of Car Characteristics",
         xaxis = list(title = ""),
         yaxis = list(title = ""))
```

**Strong negative correlations** (blue) between weight/cylinders and fuel efficiency.

---

## Line Chart: Performance Comparison

```{r line_chart, fig.width=9, fig.height=6}
# Sort by weight for line chart
mtcars_sorted <- mtcars %>%
  arrange(wt) %>%
  mutate(car_name = rownames(mtcars)[order(wt)])

# Create separate traces for each cylinder count
plot_ly() %>%
  add_trace(data = filter(mtcars_sorted, cyl == 4),
           x = ~wt, y = ~mpg,
           type = "scatter", mode = "lines+markers",
           name = "4 Cylinders",
           line = list(width = 3),
           marker = list(size = 8)) %>%
  add_trace(data = filter(mtcars_sorted, cyl == 6),
           x = ~wt, y = ~mpg,
           type = "scatter", mode = "lines+markers",
           name = "6 Cylinders",
           line = list(width = 3),
           marker = list(size = 8)) %>%
  add_trace(data = filter(mtcars_sorted, cyl == 8),
           x = ~wt, y = ~mpg,
           type = "scatter", mode = "lines+markers",
           name = "8 Cylinders",
           line = list(width = 3),
           marker = list(size = 8)) %>%
  layout(title = "MPG Trends Across Weight Range",
         xaxis = list(title = "Weight (1000 lbs)"),
         yaxis = list(title = "Miles per Gallon"),
         hovermode = "closest")
```

---

## Animated Bubble Chart

```{r animated_plot, fig.width=9, fig.height=6}
# Create time series effect by ranking
mtcars_anim <- mtcars %>%
  mutate(car = rownames(mtcars),
         frame = ntile(hp, 5))  # Create 5 frames based on HP

plot_ly(mtcars_anim,
        x = ~wt,
        y = ~mpg,
        size = ~hp,
        color = ~factor(cyl),
        frame = ~frame,
        text = ~car,
        hoverinfo = "text",
        type = "scatter",
        mode = "markers",
        marker = list(sizemode = "diameter",
                     opacity = 0.7)) %>%
  layout(title = "Cars Appearing by Horsepower Range",
         xaxis = list(title = "Weight (1000 lbs)", range = c(1, 6)),
         yaxis = list(title = "MPG", range = c(10, 35))) %>%
  animation_opts(frame = 1000, transition = 500, redraw = FALSE)
```

**Click play!** Watch how cars are distributed across the power spectrum.

---

## Key Takeaways

**What we learned:**

1. ✅ **Weight matters**: Heavier cars consume more fuel
2. ✅ **Cylinders count**: More cylinders = less efficiency
3. ✅ **Power trade-off**: High horsepower typically means lower MPG
4. ✅ **Interactive is better**: Plotly makes data exploration intuitive

**Why use Plotly?**

- Professional-looking plots with minimal code
- Interactivity engages your audience
- Web-ready and shareable
- Works seamlessly with R Markdown

---

## Resources & Next Steps

**Learn More:**

- 📚 [Official Plotly R Documentation](https://plotly.com/r/)
- 📊 [Plotly R Graphing Library](https://plotly-r.com/)
- 💡 [R Graph Gallery](https://www.r-graph-gallery.com/)

**Try it yourself:**

```r
# Basic example
library(plotly)
plot_ly(data = iris,
        x = ~Sepal.Length,
        y = ~Petal.Length,
        color = ~Species)
```

**Publishing options:**
- GitHub Pages (free hosting)
- RPubs (easy one-click publishing)
- NeoCities (simple hosting)

---

## Thank You!

<div style="text-align: center; padding: 50px;">

**Presentation created on:** `r format(Sys.Date(), "%B %d, %Y")`

**Tools used:**
- R Markdown
- Plotly for R
- Slidy presentation format

**All plots are interactive - hover, click, zoom, and explore!**

---

*Questions or feedback?*  
*[Your Email] | [Your GitHub]*

</div>
